A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-20 tex/body.tex
more

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}= \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,k_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
abbreviate
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(x), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(x)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

Given a signature $\Sig$, $\thD$ refers to the \emph{theory of
datatypes and codatatypes},
which in addition to $\Sig$ defines a class of $\Sig$-interpretations $\JJJJ$,
namely the ones that satisfy the %axioms --- existence of all infinite codatatype values cannot be axiomatized
properties mentioned in this section, including (co)induction. In particular, the interpretations
in~$\JJJJ$ share the same fixed interpretation for constructor terms and correctly
applied selector terms but may differ on wrongly applied
selector terms and on variables.
A formula $\varphi$ is $\thD$-\emph{satisfiable} if there
exists an interpretation in $\JJJJ$ that satisfies~$\varphi$.
% otherwise, it is $\thD$-unsatisfiable. (I trust the readers' intelligence when
% it comes to English prefixes like non-, in-, un-, and a-.)

@@

%\ourparagraph{Special Cases.}
{\looseness=-1
Some codatatypes are so degenerate as to have infinite values
even though they are finite. A simple example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
unique value is $\MU a.\; \const{A}(a)$. Other specimens are
$\ty{stream}_{\,\ty{unit}\,}$
and
\begin{conf}%
\vthinspace$\keyw{codatatype}~\ty{b} = \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit})
~\keyw{and}~ \ty{c} = \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})$,
\end{conf}%
\begin{rep}%
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
\end{rep}%
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
crucial to detect these. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to make \ty{unit} an ordinary type, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, \begin{rep}given cardinalities for the ordinary types, \end{rep}%
it is easy to characterize this degenerate case.

>>>>

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}= \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
abbreviate
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

Given a signature $\Sig$, $\thD$ refers to the \emph{theory of
datatypes and codatatypes},
which %in addition to $\Sig$
defines a class of $\Sig$-interpretations $\JJJJ$,
namely the ones that satisfy the %axioms --- existence of all infinite codatatype values cannot be axiomatized
properties mentioned in this section, including (co)induction. The interpretations
in~$\JJJJ$ share the same interpretation for constructor terms and correctly
applied selector terms (modulo renaming of domain elements) but may differ on
variables and wrongly applied selector terms.
A formula is $\thD$-\emph{satisfiable} if there
exists an interpretation in $\JJJJ$ that satisfies it.
% otherwise, it is $\thD$-unsatisfiable. (I trust the readers' intelligence when
% it comes to English prefixes like non-, in-, un-, and a-.)

@@

%\ourparagraph{Special Cases.}
{\looseness=-1
Some codatatypes are so degenerate as be finite even though they
have infinite values. A simple example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
unique value is $\MU a.\; \const{A}(a)$. Other specimens are
$\ty{stream}_{\,\ty{unit}\,}$
and
\begin{conf}%
\vthinspace$\keyw{codatatype}~\ty{b} = \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit})
~\keyw{and}~ \ty{c} = \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})$,
\end{conf}%
\begin{rep}%
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
\end{rep}%
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
crucial to detect these. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to make \ty{unit} an ordinary type, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, \begin{rep}given cardinalities for the ordinary types, \end{rep}%
it is easy to characterize this degenerate case.
